Effect of Black Hole Spins on Christodoulou Memory in Binary Mergers and its implication on Observing Gravitational Wave Bursts With Memory in Pulsar Timing Array

### 2019-10-10 notes/Paper.tex
Updated notes

\par The corrections to the leading-oder formula for the nonlinear memory's contribution to 3PN order for non-spinning binaries has been computed by by Favata \cite{Favata2009}. However the effect of binary spins have not been explored in these results. Astrophysical black holes are formed by gravitation collapse of massive star when when they run out of their nuclear fuel in the core. These massive stars are likely to posses spin angular momentum, as a result the black hole formed would also be spinning. The black holes can also accrete matter from surrounding and as a result spin up during the process. It is highly likely that there is a large population of spinning binaries in the universe.      
\par In this paper we calculate the nonlinear memory contribution to the + waveform polarization and look at the effect of spin contribution. For PN calculations, this paper is just an extension of Favata's\cite{Favata2009} paper, where we include the spin contribution up to leading (1.5 PN) order for nonprecessing binaries. The nonlinear memory contribution near merger is computed using NR data from publicly available SXS catalog, where we process the data to remove the low frequency numerical noise. In Sec. II, we describe the calculation required to compute memory contribution to the post-Newtonian waveform of quasi-circular, inspiralling binaries. In Sec. III, we describe the method used to extract memory contribution from NR simulations. Sec. IV we give the results which looks at the effect of spin on total nonlinear memory. We use PN expressions for memory for early time and for late time we use memory extracted from NR simulations. In Sec. V we look at how these results can effect expected results in effort to detect the BWM from super massive binaries in pulsar timing array. We look a the timing residuals growth over a period of two weeks interval of time.

@@

\begin{figure}
 	\includegraphics[width=3.0in]{../plots/MemoryPlot_nonSpining/q7.pdf}
 	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. This figure shows a comparison for different mass ratios. It can be seen that higher mass ration tends to decrease the Chistodoulou memory. The early inspiral part is computed using post-Newtonian expression and the later part during BBH merger is computer using NR data. The red dashed line shows result from PN expression}
 	\label{fig:differentmassratio}
\end{figure}
\begin{figure}
	\includegraphics[width=3.0in]{../plots/MemoryPlot_AlignedSpinSXSdata/0p99.pdf}
	\includegraphics[width=3.0in]{../plots/MemoryPlot_AntialignedSpinSXSdata/m0p94.pdf}
	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. The top figure shows a comparison for different values of dimensionless spin parameter $\chi$. These results are for BBH system with equal spin magnitude for individual black hole and spins aligned with orbital angular momentum. It can be seen that higher spin magnitude tend to increase the Christodoulou memory. The bottom figure shows the anti aligned case.  }
	\label{fig:differntchivalues}
\end{figure}
\section{Observing Memory from merger of Super-massive black holes  }
Nonlinear memory from the merger of super massive black hole binary can be potentially observed in pulsar timing measurement. Gravitational waves, when they pass through the space between earth and pulsar effect the time of arrival of the radio pulses at earth. The modulation in the observed frequency is given by
\begin{equation}
\frac{\Delta \nu}{\nu_{0}} = \frac{1}{2c}\int_{0}^{D}d\lambda \left(e^{i}e^{j}\frac{\partial h_{ij}}{\partial t}\right) \Bigg\vert_{path}
\end{equation}
The pulsar timing measurement primarily measures the difference between the actual pulse arrival time and the time predicted from the spin-down model of a pulsar, which is called the timing residuals. The timing residuals $s(t)$, accumulated during a time interval of length due to passing gravitational waves is given by the following expression 
\begin{equation}
	s(t) = \frac{1}{2}(1+\mu)\left[\left(\int_{0}^{t}d\tau \, h_{+}(\tau)\right)\cos 2\phi + \left(\int_{0}^{t}d\tau \, h_{\times}(\tau)\right)\sin 2\phi\right]
\end{equation}
There are two unknown quantities associated with the timing residuals, the pulsar periods and it's spin down rate. We subtract these two quantities from the expression $s(t)$.

>>>>

\par The corrections to the leading-oder formula for the nonlinear memory's contribution to 3PN order for non-spinning binaries has been computed by by Favata \cite{Favata2009}. However the effect of binary spins have not been explored in these results. Astrophysical black holes are formed by gravitation collapse of massive star when when they run out of their nuclear fuel in the core. These massive stars are likely to posses spin angular momentum, as a result the black hole formed would also be spinning. The super black holes found at the galactic center can acquire significant spins due also accretion of matter from surrounding. It is highly likely that there is a large population of spinning super massive binaries in the universe.      
\par In this paper we calculate the nonlinear memory contribution to the + waveform polarization and look at the effect of spin contribution. For PN calculations, we extend the calculations  done in Favata's\cite{Favata2009} paper, where we include the spin contribution up to leading (1.5 PN) order for nonprecessing binaries. The nonlinear memory contribution near merger is computed using NR data from publicly available SXS catalog, where we process the data to remove the low frequency numerical noise. In Sec. II, we describe the calculation required to compute memory contribution to the post-Newtonian waveform of quasi-circular, inspiralling binaries. In Sec. III, we describe the method used to extract memory contribution from NR simulations. Sec. IV we give the results which looks at the effect of spin on total nonlinear memory. We use PN expressions for memory for early time and for late time we use memory extracted from NR simulations. In Sec. V we look at how these results can effect expected results in effort to detect the BWM from super massive binaries in pulsar timing array. We look a the timing residuals growth over a period of two weeks interval of time.

@@

\begin{figure}
 	\includegraphics[width=3.5in]{../plots/MemoryPlot_nonSpining/q7.pdf}
 	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. This figure shows a comparison for different mass ratios. It can be seen that higher mass ration tends to decrease the Chistodoulou memory. The early inspiral part is computed using post-Newtonian expression and the later part during BBH merger is computer using NR data. The red dashed line shows result from PN expression}
 	\label{fig:differentmassratio}
\end{figure}
\begin{figure}
	\includegraphics[width=3.5in]{../plots/MemoryPlot_AlignedSpinSXSdata/0p99.pdf}
	\includegraphics[width=3.5in]{../plots/MemoryPlot_AntialignedSpinSXSdata/m0p94.pdf}
	\caption{The figure above shows Christodoulou memory as a function of time during BBH coalesce. The top figure shows a comparison for different values of dimensionless spin parameter $\chi$. These results are for BBH system with equal spin magnitude for individual black hole and spins aligned with orbital angular momentum. It can be seen that higher spin magnitude tend to increase the Christodoulou memory. The bottom figure shows the anti aligned case.  }
	\label{fig:differntchivalues}
\end{figure}
\section{Observing Memory from merger of Super-massive black holes  }
Nonlinear memory from the merger of super massive black hole binary can be potentially observed in pulsar timing measurement. Gravitational waves, when they pass through the space between earth and pulsar effect the time of arrival of the radio pulses at earth. The modulation in the observed frequency is given by
\begin{equation}
\frac{\Delta \nu}{\nu_{0}} = \frac{1}{2c}\int_{0}^{D}d\lambda \left(e^{i}e^{j}\frac{\partial h_{ij}}{\partial t}\right) \Bigg\vert_{path}
\end{equation}
The pulsar timing measurement primarily measures the difference between the actual pulse arrival time and the time predicted from the spin-down model of a pulsar, which is called the timing residuals. The timing residuals $s(t)$, accumulated during a time interval of length due to passing gravitational waves is given by the following expression 
\begin{equation}
	s(t) = \frac{1}{2}(1+\mu)\left[\left(\int_{0}^{t}d\tau \, h_{+}(\tau)\right)\cos 2\phi + \left(\int_{0}^{t}d\tau \, h_{\times}(\tau)\right)\sin 2\phi\right]
\end{equation}
There are two unknown quantities associated with the timing residuals, the pulsar periods and it's spin down rate. We subtract these two quantities from the expression $s(t)$.
